---
title: "Mini Literate Analysis"
author: "Shayna Sura"
format: html
---

## Introduction
### About this data
This data came from the [Arctic Data Center](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X) and contains seawater chemistry data from the Arctic Ocean near the North Pole collected from 2000 - 2014. The data was retrieved on 2 October 2023.

### Setup

```{r}
#| echo: false

library(readr)
library(here)

# the here function from the here package sets the default directory to wherever your Rproject lives.
```

### Read data
```{r}
#| warning: false

# bg_chem <- read_csv("../data/BGchem2008data.csv")

bg_chem <- read_csv(here("data/BGchem2008data.csv"))
# the here() function from the here package sets the default directory to wherever your Rproject lives.
# the here() function really helps with reproducibility.

```

Exploring our data
```{r}
#| eval: false

colnames(bg_chem)
str(bg_chem)

summary(bg_chem)

unique(bg_chem$Date)
#will return the unique values from a column

# View(bg_chem)

```



## Analysis

### Calculate summary statistics
```{r}

nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)

```


### Calculate mean of Redfield ratio
```{r}

ratio <- (nitrate+nitrite+amm)/phos

# round(ratio)

```


### Plot Redfield ratio
```{r}
plot(bg_chem$P, bg_chem$NO2+bg_chem$NO3+bg_chem$NH4)
```


## Conclusion

You can access variables from your environment in your text! Example:

The Redfield ratio of this data is `r round(ratio, digits=2)` or `r ratio`.


